Transient Corotating Clumps Around Adolescent Low-Mass Stars: Four Years of Complex Rotators from TESS

### 2023-08-29 paper/ms.tex
Update ms; add completeness figure.
More clear than trying to explain it in words!

Co-authored-by: Joshua Winn <jnwinn@princeton.edu>

\received{---}
\revised{---}
\accepted{---}
\shorttitle{Four Years of CRs}

@@

% 250 word max!
\begin{abstract}
  %Complex rotators (CRs) are low-mass
  %pre-main-sequence stars with nearly periodic optical modulation.
  %The modulation is probably caused by eclipsing clumps of dust or gas
  %in orbit at the Keplerian corotation radius.  Here, we report new
  %CRs discovered in TESS short-cadence data collected between
  %2018~July and 2022~September.  Our search of \nstarssearched\ K and
  %M dwarfs with $T$$<$16 and $d$$<$150\,pc yielded \ngoods\
  %high-quality CRs.  Most of these discoveries are new, and they
  %include the brightest ($T$$\approx$9.5), closest
  %($d$$\approx$20\,pc), and oldest ($\approx$200\,Myr) examples of
  %this class of object currently known.  A few objects are outliers:
  %LP 12-502 for instance showed a ``dip complex'' with a period and
  %duration that were fixed over 1,500 cycles, but in detail, this
  %system exhibited between four and eight local minima per cycle, and
  %at times also displayed multiple periods simultaneously.  We
  %demonstrate that transient corotating material is the most viable
  %explanation for this class of object, and present evidence
  %supporting the hypothesis that this material is sculpted by stellar
  %magnetic fields dominated by high order multipoles.  We expect that
  %our sample will facilitate future efforts aimed at connecting CRs
  %to the broader contexts of star, disk, and perhaps even exoplanet
  %evolution. 	
  % % JNW alternate:
  Complex rotators (CRs) are low-mass pre-main-sequence stars with
  light curves characterized by periods of 0.2--2 days, amplitudes of
  a few percent, and slowly-evolving sharp features.  The sharp
  features might be caused by eclipsing clumps of dust or gas orbiting
  at the corotation radius.  To improve our understanding of these
  enigmatic stars, we conducted a survey for CRs using TESS
  short-cadence data collected between 2018~July and 2022~September.
  Our search of \nstarssearched\ K and M dwarfs with $T$$<$16 and
  $d$$<$150\,pc yielded \ngoods\ high-quality CRs, most of which are
  new discoveries.  Our sample includes the brightest
  ($T$$\approx$9.5), closest ($d$$\approx$20\,pc), and oldest
  ($\approx$200\,Myr) known CRs.  One of the new discoveries, LP
  12-502, stands out for the complexity of its light curve --- with as
  many as eight flux dips per cycle --- and the detail with which the
  evolution in each dip's amplitude and phase can be tracked over
  1{,}500 cycles.  We argue that corotating gas or dust remains the
  most viable explanation for CRs. The gas or dust is probably
  entrained by the star's magnetic field, and the sharp features may
  result from a multipolar field topology, a hypothesis supported by
  correspondences between the light curves of CRs and of two B stars
  known to have strong multipolar magnetic fields.
\end{abstract}

@@

Figure~\ref{fig:f1} illustrates two proposed models for the extrinsic
material.  The first scenario invokes opaque dust ``clumps'' that
orbit near the Keplerian corotation radius [$R_{\rm cr} =
(GM/\Omega^2)^{1/3}$, where $\Omega = 2\pi/P_{\rm rot}$] and
periodically transit the star
\citep{2017AJ....153..152S,2023MNRAS.518.4734S}.  The second scenario
invokes ``prominences'', long-lived condensations of cool, dense,
marginally-ionized gas that are embedded within the hotter corona and
corotate with the star
\citep{1989MNRAS.238..657C,2019MNRAS.482.2853J,2022MNRAS.514.5465W}.
These hypothetical structures are analogous to quiescent prominences
and filaments seen in the solar corona \citep[see
e.g.][]{2015ASSL..415.....V}, though rather than existing at a
fraction of the stellar radius as in the solar case, they would exist
at distances of a few stellar radii.  A final possibility is that a
thin optically-thick ring obscures a small portion of the stellar
photosphere \citep{2019ApJ...876..127Z}.  Hot spots passing behind
such a ring could produce sudden dips.  We do not favor this scenario
due to two issues discussed in Appendix~\ref{app:ring}.

@@

		\vspace{-0.6cm}
		\subfloat{
			\includegraphics[width=0.66\textwidth]{f1b.pdf}
		}
	\end{center}
		\vspace{-0.3cm}
	\caption{
		{\bf Complex rotators (CRs)}:
    {\it Top:} Phase-folded TESS light curves of three CRs.  Each panel shows
    the average of the data accumulated over one month.
    Gray circles are raw 2-minute data; black
    circles are binned to 300 points per cycle.  The period in hours
    is printed in the
    bottom right corner.  Left-to-right, the objects are LP
    12-502 (TIC 402980664; Sector~19), TIC 94088626 (Sector 10), and
    TIC 425933644 (Sector~28).
    {\it Bottom:} Cartoon explanations for the phenomenon.  The dust
    clump scenario (left) and prominence scenario (right) both invoke
    magnetically-confined corotating material.
	}
	\label{fig:f1}
\end{figure*}

@@

To construct our sample, we analyzed the short-cadence data acquired
by TESS between 2018 July 25 and 2022 September 1 (Sectors 1-55).
Specifically, we used the 120-second cadence light curves produced by
the Science Processing and Operations Center at NASA Ames
\citep{2016SPIE.9913E..3EJ}.  While the TESS data products from these
sectors also included full frame images with cadences of 600 and 1800
seconds, we limited our scope in this study for the sake of simplicity
in data handling and uniformity in the data.  In exchange, we
sacrificed both completeness and homogeneity of the selection
function.  While TESS cumulatively observed $\approx$90\% of the sky
for at least one lunar month between 2018 July and 2022 September, the
120-second cadence data were collected for only a subset of observable
stars during this time due to telemetric constraints
\citep[see][]{2021PASP..133i5002F}.

To assess the completeness of the resulting 120-second cadence data
that are the basis of this study, we cross-matched TIC8.2
\citep{2019AJ....158..138S,2021arXiv210804778P} against the Gaia DR2 point-source catalog
\citep{2018A&A...616A...1G}.  We opted for Gaia DR2 rather than DR3
because the base catalog for TIC8 was Gaia DR2, which facilitated a
one-to-one crossmatch using the Gaia source identifiers.  This
exercise showed that for $T$$<$16 M dwarfs, the TESS 2-minute data are
$\approx$50\% complete at $\approx$50\,pc.  At $<$20\,pc,
$\gtrsim$80\% of the $T$$<$16 M dwarfs have at least one sector of
short-cadence data; at $>$100\,pc, $\lesssim$10\% of such M dwarfs
have at least one sector of short-cadence data.  Armed with this
understanding, we then used our cross-match between Gaia DR2 and TIC8
to select our stars of interest, which we defined as stars with
120-second cadence TESS light curves satisfying these four conditions:
\begin{align}
  T &< 16 \quad&(\mathrm{Amenable\ with\ TESS})\\
  \bprp &> 1.5 \quad&(\mathrm{Red\ stars\ only})\\
  M_{\rm G} &> 4 \quad&(\mathrm{Dwarf\ stars\ only})\\
  d &< 150\,{\rm pc} \quad&(\mathrm{Close\ stars\ only}).
\end{align}
Here, $M_{\rm G} = G + 5\log(\varpi_{\rm as}) + 5$ is the Gaia
$G$-band absolute magnitude, $\varpi_{\rm as}$ is the parallax in
units of arcseconds, and $d$ is a geometric distance defined by
inverting the parallax and ignoring any zero-point correction.  The
target sample therefore includes \nstarssearched\ M dwarfs and late-K
dwarfs, down to $T$$<$16 and out to $d$$<$150\,pc.

>>>>

\received{---}
\revised{---}
\accepted{---}
\shorttitle{Four Years of Complex Rotators}

@@

% 250 word max!
\begin{abstract}
  Complex rotators (CRs) are low-mass pre-main-sequence stars with
  light curves characterized by periods of 0.2--2 days, amplitudes of
  a few percent, and slowly-evolving sharp features.  The sharp
  features might be caused by eclipsing clumps of dust or gas orbiting
  at the corotation radius.  To improve our understanding of these
  enigmatic stars, we conducted a survey for CRs using TESS
  short-cadence data collected between 2018~July and 2022~September.
  Our search of \nstarssearched\ K and M dwarfs with $T$$<$16 and
  $d$$<$150\,pc yielded \ngoods\ high-quality CRs, most of which are
  new discoveries.  Our sample includes the brightest
  ($T$$\approx$9.5), closest ($d$$\approx$20\,pc), and oldest
  ($\approx$200\,Myr) known CRs.  One of the new discoveries, LP
  12-502, stands out for the complexity of its light curve --- with as
  many as eight flux dips per cycle --- and the detail with which the
  evolution in each dip's amplitude and phase can be tracked over
  1{,}500 cycles.  We argue that corotating gas or dust remains the
  most viable explanation for CRs. The gas or dust is probably
  entrained by the star's magnetic field, and the sharp features may
  result from a multipolar field topology, a hypothesis supported by
  correspondences between the light curves of CRs and of two B stars
  known to have strong multipolar magnetic fields.
\end{abstract}

@@

Figure~\ref{fig:f1} illustrates two proposed models for the extrinsic
material.  The first scenario invokes opaque dust ``clumps'' that
orbit near the Keplerian corotation radius [$R_{\rm c} =
(GM/\Omega^2)^{1/3}$, where $\Omega = 2\pi/P_{\rm rot}$] and
periodically transit the star
\citep{2017AJ....153..152S,2023MNRAS.518.4734S}.  The second scenario
invokes ``prominences'', long-lived condensations of cool, dense,
marginally-ionized gas that are embedded within the hotter corona and
corotate with the star
\citep{1989MNRAS.238..657C,2019MNRAS.482.2853J,2022MNRAS.514.5465W}.
These hypothetical structures are analogous to quiescent prominences
and filaments seen in the solar corona \citep[see
e.g.][]{2015ASSL..415.....V}, though rather than existing at a
fraction of the stellar radius as in the solar case, they would exist
at distances of a few stellar radii.  A final possibility is that a
thin optically-thick ring obscures a small portion of the stellar
photosphere \citep{2019ApJ...876..127Z}.  Hot spots passing behind
such a ring could produce sudden dips.  We do not favor this scenario
due to two issues discussed in Appendix~\ref{app:ring}.

@@

		\vspace{-0.6cm}
		\subfloat{
			\includegraphics[width=0.66\textwidth]{f1b.pdf}
		}
	\end{center}
		\vspace{-0.3cm}
	\caption{
		{\bf Complex rotators (CRs)}:
    {\it Top:} Phase-folded TESS light curves of three CRs.  Each
    panel shows the average of the data accumulated over one month.
    Gray circles are raw 2-minute data; black circles are binned to
    300 points per cycle.  The period in hours is printed in the
    bottom right corner.  Left-to-right, the objects are LP 12-502
    (TIC 402980664; Sector~19), TIC 94088626 (Sector 10), and TIC
    425933644 (Sector~28).
    {\it Bottom:} Cartoon explanations for the phenomenon.  The dust
    clump scenario (left) and prominence scenario (right) both invoke
    magnetically-entrained corotating material.
	}
	\label{fig:f1}
\end{figure*}

@@

To construct our sample, we analyzed the short-cadence data acquired
by TESS between 2018 July 25 and 2022 September 1 (Sectors 1-55).
Specifically, we used the 120-second cadence light curves produced by
the Science Processing and Operations Center at the NASA Ames Research
Center \citep{2016SPIE.9913E..3EJ}.  While the TESS data products from
these sectors also included full frame images with cadences of 600 and
1800 seconds for a larger number of sources, we restricted our
attention to the 120-second data for the sake of uniformity and
simplicity in data handling.  In exchange, we sacrificed both
completeness and homogeneity of the selection function.  While TESS
cumulatively observed $\approx$90\% of the sky for at least one lunar
month between 2018 July and 2022 September, the 120-second cadence
data were collected for only a subset of observable stars that were
preferentially nearby and bright \citep[see][]{2021PASP..133i5002F}.
The total 120-second data volume
from Sectors 1-55 included 1{,}087{,}475 short-cadence light curves, which were
available for 428{,}121 unique stars.
% latter stats from MERGED_spoc2min_sector1_to_sector58.csv

To further simplify our search, we defined our target sample as
stars with 120-second cadence TESS light curves satisfying the
following four conditions:
\begin{align}
  T &< 16 \quad&(\mathrm{Amenable\ with\ TESS}) \label{eq:one}\\
  \bprp &> 1.5 \quad&(\mathrm{Red\ stars\ only})\\
  M_{\rm G} &> 4 \quad&(\mathrm{Dwarf\ stars\ only})\\
  d &< 150\,{\rm pc} \quad&(\mathrm{Close\ stars\ only}) \label{eq:four}.
\end{align}
Here, $M_{\rm G} = G + 5\log(\varpi_{\rm as}) + 5$ is the Gaia
$G$-band absolute magnitude, $\varpi_{\rm as}$ is the parallax in
units of arcseconds, and $d$ is a geometric distance defined by
inverting the parallax and ignoring any zero-point correction.  We
performed this selection by cross-matching TIC8.2
\citep{2019AJ....158..138S,2021arXiv210804778P} against the Gaia DR2
point-source catalog \citep{2018A&A...616A...1G}.  We opted for Gaia
DR2 rather than DR3 because the base catalog for TIC8 was Gaia DR2,
which facilitated a one-to-one crossmatch using the Gaia source
identifiers.  The target sample ultimately included \nstarssearched\ M
dwarfs and late-K dwarfs, down to $T$$<$16 and out to $d$$<$150\,pc.
For stars with multiple sectors of TESS data available, we searched
for CR signals independently.
In total, our \nstarssearched\ star target list included \nlcssearched\ month-long light
curves.

We assessed the completeness of our selection function by comparing
the number of stars with TESS Sector 1-55 short-cadence data against
the number of Gaia DR2 point sources.
We required all stars to meet conditions~\ref{eq:one}--\ref{eq:four}.
The results are shown in Figure~\ref{fig:completeness}.
TESS 2-minute data exist for $\approx$50\%
of $T$$<$16 M and late-K dwarfs at $\approx$50\,pc.
Within 20\,pc, $\gtrsim$80\% of the $T$$<$16 M and late-K dwarfs have
at least one sector of short-cadence data.
Beyond 100\,pc, $\lesssim$10\% of such stars have any short-cadence
data available.
This can be translated into our sensitivity for the lowest mass stars
by considering that the spectral type of a $T$=16 star at $d$=50\,pc
is $\approx$M5.5V, corresponding to a main-sequence mass of
$\approx$0.12\,$M_\odot$.

\begin{figure}[!tp]
	\begin{center}
		\centering
		\includegraphics[width=0.48\textwidth]{f2completeness.pdf}
		\vspace{-0.5cm}
		\caption{
			{\bf Completeness of the TESS 2-minute cadence data for late-K
      and early M dwarfs near the Sun, from Sectors 1-55.}  
      The orange dotted curve shows the number of stars in successive
      radial shells, each with a width of 10\,pc.  To be counted, these stars must meet
      the listed conditions: they must
      be red ($\bprp$>$1.5$)
      dwarf stars ($M_{\rm G}$>4) amenable for TESS observations ($T$<16).
      The blue solid curve shows the
      fraction of such stars with at least one sector of TESS
      2-minute cadence data acquired between Sectors 1-55.  The
      latter data comprise the selection function for our complex
      rotator search.
		}
			\vspace{-0.5cm}
		\label{fig:completeness}
	\end{center}
\end{figure}
